---
title: 'Interaction Analysis small RNA'
author: "Franco"
date: "12/10/2022"
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: kate
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes
  pdf_document:
    latex_engine: lualatex
    pandoc_args: --listings
    toc: yes
subtitle: "Subsampled to 700 000"
---

```{r, message=FALSE, warning=FALSE}
require(nlme)
require(reshape2)
require(ggplot2)
require(multcomp)
require(emmeans)
require(rstatix)
require(ggpubr)
require(RColorBrewer)
require(tidyverse)
require(dplyr)
require(ggbeeswarm)
require(readxl)
require(broom)
require(scales)
require(RCurl)
require(plotly)
require(ggrepel)
require(gridExtra)
require(ggExtra)
require(DT)
source("./HelperFunctions.R")


theme_point<-theme_bw()+theme(strip.background = element_blank())
theme_bar<-theme_bw()+theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),strip.background = element_blank(),axis.ticks.x = element_blank(),axis.line.x = element_blank())
theme_boxplot<-theme_classic()+theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1),strip.background = element_blank(),axis.ticks.x = element_blank(),axis.title.x = element_blank(),axis.line.x = element_blank(),legend.position = "none")
color_panel<-c("#e35d6a","#5bb75b","#428bca","#e87810","#23496b","#ffbf00","#cc2028","#039748","pink","darkgray")
names(color_panel)<-sort(c("serum","Biomatrica","EDTA separator","EDTA","Citrate","RNA Streck","Roche","DNA Streck", "PAXgene", "ACD-A"))

color_panel1 <-  c("#039748","#039748","#ffbf00","#ffbf00","#e35d6a","#e35d6a","#5bb75b","#5bb75b","#428bca","#428bca","#23496b","#23496b","#cc2028","#cc2028","#e87810")
color_panel2 <-  brewer.pal(10, name = "Paired")
names(color_panel2) <- c("serum","Biomatrica","EDTA separator","EDTA","Citrate","RNA Streck","Roche","DNA Streck", "PAXgene", "ACD-A")
full_nr <- format_format(big.mark = ",", decimal.mark = ".", scientific = FALSE)


color_panel_spikes <- c("#e35d6a","gray","#5bb75b","#428bca")
names(color_panel_spikes) <- c("LP1","MIMAT","RC1","RC2")
color_panel_spikes2 <- c("#23496b","#ffbf00","gray","#ffbf01","orange","#23496b")
names(color_panel_spikes2) <- c("LP","RC","MIMAT","RC1","RC2","LP1")

set.seed(1)
```

# Annotation

```{r, message=FALSE, warning=FALSE}
#Annotation
sample_annotation<-read_tsv("./Annotation_exRNAQC017.csv")
colnames(sample_annotation)[23]<-'TimeInterval' 
sample_annotation<-sample_annotation %>% mutate(donorID= case_when((Donor == "PNL-XNID") ~ "D5",
                                                                   (Donor == "PNL-7DEN") ~ "D1",
                                                                   (Donor == "PNL-8ZI1") ~ "D2",
                                                                   (Donor == "PNL-NLID") ~ "D3",
                                                                   (Donor == "PNL-UCH7") ~ "D4"), 
                                                SampleID= paste0(GraphKit,"_",GraphTube,"_",TimeInterval,"_",donorID),
                                                tube_kitID=paste0(GraphTube,"_",GraphKit),
                                                tube_kit_timeID=paste0(GraphTube,"_",GraphKit,"_",TimeInterval),
                                                PlasmaInputVml= PlasmaInputV/1000)
    
sample_annotation$Tube<-as.factor(sample_annotation$Tube)
sample_annotation$RNAisolation<-as.factor(sample_annotation$RNAisolation)
sample_annotation$TimeInterval<-as.factor(sample_annotation$TimeInterval)
sample_annotation$donorID<-as.factor(sample_annotation$donorID)

#Reads
miRNAs <- data.table::fread("mergedTxts/allmiRs_subs.txt", header=T, data.table = F)
spikes <- data.table::fread("mergedTxts/allspikes_subs.txt", header=T, data.table = F)
reads <- data.table::fread("mergedTxts/allreads_subs.txt", header=T, data.table = F)
contam <- data.table::fread("mergedTxts/allcontam_subs.txt", header=T, data.table = F)

reads_total<- read_tsv("./smallRNA_table.tsv")
reads_total <- full_join(reads_total, sample_annotation, by = "RNAID")

Tube.levels<-levels(reads_total$Tube)
RNAisolation.levels<-levels(reads_total$RNAisolation)
TimeInterval.levels<-levels(reads_total$TimeInterval)

#ggplot(reads_total, aes(x = TimeInterval, y = total_reads, group = donorID:RNAisolation)) +
#  geom_line(aes(color=RNAisolation))+geom_point() + facet_grid(. ~ Tube) + labs(title = "Raw Number of Reads")

spikes_sum <- spikes %>% gather(key="UniqueID",value="spikecounts",-"spikeID") %>% 
  group_by(UniqueID) %>% dplyr::summarise(counts=sum(spikecounts, na.rm=T)) %>%
  #spread(key=UniqueID,value=spikesum) %>% 
  mutate(reads="spikes")

# add sum of spikes to the total mapped reads (these were not considered as "mapped" yet)
reads_complete <- reads %>% gather(key="UniqueID",value="counts",-"reads") %>% 
  rbind(., spikes_sum) %>%
  spread(key=reads, value=counts) %>% 
  mutate(all_mapped = mapped + spikes) %>% 
  gather(key="reads",value="counts",-"UniqueID")

DT::datatable(sample_annotation %>% dplyr::select(c("RNAID","RNAisolation","PlasmaInputV","donorID","Tube","TimeInterval","Sequinconc","ERCCconc",Abbrevation="GraphKit")), rownames = TRUE, filter="top", options = list(pageLength = 10, scrollX=T), caption = "Sample annotation table")
```

# RNA concentration

## Endogenous vs LP spikes: relative small RNA concentration in eluate

calculation: count mapped miRNAS / count mapped LP spikes

```{r, message=FALSE, warning=FALSE}
gene_level_ratios <- rbind(
  reads %>% 
  filter(reads=="mapped_miR") %>%
  mutate(type="endogenous") %>%
    group_by(type) %>%
    dplyr::summarise_if(is.numeric, sum, na.rm=TRUE),spikes %>%
    mutate(type=gsub(".-..$","",spikeID)) %>%
    group_by(type) %>%
    dplyr::summarise_if(is.numeric, sum, na.rm=TRUE)
  ) %>%
  gather(., key="UniqueID",value="counts",-type) %>%  #long format
  spread(., key = "type", value="counts") %>%
  mutate("LPvsEndo"=LP/endogenous, "RCvsEndo"=RC/endogenous, "EndovsRC"=endogenous/RC, "EndovsLP"= endogenous/LP) %>%
  left_join(., sample_annotation %>%
              dplyr::select(c("UniqueID","RNAID","RNAisolation","Tube","SampleID","GraphTube","EluateV","PlasmaInputV", "BiologicalReplicate", "TimeInterval","Donor", "donorID")), by=c("UniqueID" = "RNAID"))  #add annotation
```

```{r}
gene_level_ratios <- gene_level_ratios %>%
  mutate("EndovsLP_InputCorr"= (EndovsLP)*EluateV/PlasmaInputV)

Tube.levels<-levels(gene_level_ratios$Tube)
RNAisolation.levels<-levels(gene_level_ratios$RNAisolation)
TimeInterval.levels<-levels(gene_level_ratios$TimeInterval)
```

Linear mixed-effects model with crossed fixed effects of tube, kit and TimeInterval. 

```{r}
EndovsLP_m1<-lme(EndovsLP~Tube*RNAisolation*TimeInterval,
       random=~1|donorID, 
       weights = varIdent(form = ~ 1 | RNAisolation),
       data=gene_level_ratios)

anova(EndovsLP_m1)


EndovsLP_m1<-lme(EndovsLP~Tube*RNAisolation*TimeInterval,
       random=~1|donorID, 
       data=gene_level_ratios)

anova(EndovsLP_m1)

```

Only Tube, and the interactions Tube:RNAisolation and Tube:TimeInterval are significant. There are no significant three-way interactions.

Next, we look at second-order interactions:

```{r}
EndovsLP_m2<-lme(EndovsLP~(Tube+RNAisolation+TimeInterval)^2,
       random=~1|donorID, 
       data=gene_level_ratios)

anova(EndovsLP_m2)
```
```{r}
EndovsLP_m3<-lme(EndovsLP~Tube+RNAisolation+TimeInterval+Tube:RNAisolation+Tube:TimeInterval,
       random=~1|donorID, 
       data=gene_level_ratios)

anova(EndovsLP_m3)
```



```{r, include=FALSE}
#Comparing the models with and without interactions
#only significant interactions
EndovsLP_m1.Inter<-lme(EndovsLP~Tube+RNAisolation+TimeInterval+Tube:RNAisolation+Tube:TimeInterval,
       random=~1|donorID, 
       data=gene_level_ratios, method = "ML")

#no interactions
EndovsLP_m1.noInter<-lme(EndovsLP~Tube+RNAisolation+TimeInterval,
       random=~1|donorID, 
       data=gene_level_ratios, method = "ML")

#all interactions
EndovsLP_m1.allInter<-lme(EndovsLP~Tube*RNAisolation*TimeInterval,
       random=~1|donorID, 
       data=gene_level_ratios, method = "ML")

anova(EndovsLP_m1.noInter, EndovsLP_m1.allInter)
anova(EndovsLP_m1.noInter, EndovsLP_m1.Inter)
anova(EndovsLP_m1.Inter, EndovsLP_m1.allInter)
```

checking  normality of residuals

```{r}
qqnorm(EndovsLP_m3$residuals)
qqline(EndovsLP_m3$residuals)
```
```{r}
EndovsLP_m.time<-emmeans(EndovsLP_m3, ~TimeInterval|Tube)
plot(EndovsLP_m.time, comparisons=TRUE)
```
```{r}
pairs(EndovsLP_m.time)
```
Now, compare Kits

```{r}
EndovsLP_m.method<-emmeans(EndovsLP_m3, ~Tube|RNAisolation)
plot(EndovsLP_m.method, comparisons=TRUE)
```
```{r}
pairs(EndovsLP_m.method)
```

## Endogenous vs RC spikes: relative smallRNA in plasma

calculation: count mapped miRNAS / count mapped RC spikes

```{r,message=FALSE, warning=FALSE}
gene_level_ratios$GraphTube2<- paste(gene_level_ratios$Tube, gene_level_ratios$TimeInterval, sep = "_") 
test <- log_rescaling_CI(gene_level_ratios, measurevar="EndovsRC", groupvar=c("GraphTube2"))
# Plot LP/endo in log10 scale
spikes2 <- ggplot(test, aes(x=GraphTube2, y=10^measurevar_log_resc)) + 
  #geom_bar(position=position_dodge(), stat="identity")+
  geom_errorbar(aes(ymin=ci_lower_oriscale, ymax=ci_upper_oriscale), colour="grey", width=.1) +
  geom_point(aes(colour=Tube), size=1.2) +
  geom_point(data=test, aes(x=GraphTube2, y=mean_oriscale), shape=4, colour="grey") +
  geom_hline(yintercept = 1, linetype="dashed",colour="grey88") +
  facet_wrap(~RNAisolation, scales = "free_x", ncol = 2)+
  labs(x="", y="relative small RNA conc. (rescaled)", title=NULL, col = NULL, fill = NULL) +
  scale_colour_manual(values=color_panel) +
  scale_y_log10() +
  theme_bar 
ggplotly(spikes2)
```



```{r}
EndovsRC_m1<-lme(EndovsRC~Tube*RNAisolation*TimeInterval,
       random=~1|donorID, 
       data=gene_level_ratios)

anova(EndovsRC_m1)
```

```{r}
anova_EndovsRC_m1 <- round(anova(EndovsRC_m1)[c(5, 6, 7), c(4)], digits = 3)
```

```{r}
EndovsRC_m2<-lme(EndovsRC~(Tube+RNAisolation+TimeInterval)^2,
       random=~1|donorID, 
       data=gene_level_ratios)

anova(EndovsRC_m2)
```

```{r}
EndovsRC_m3<-lme(EndovsRC~Tube+RNAisolation+TimeInterval+Tube:TimeInterval,
       random=~1|donorID, 
       data=gene_level_ratios)

anova(EndovsRC_m3)
```


```{r}
EndovsRC.time<-emmeans(EndovsRC_m3, ~TimeInterval|Tube)
plot(EndovsRC.time, comparisons=TRUE)
```

```{r}
pairs(EndovsRC.time)
```

EDTA has a significantly higher smallRNA concentration in plasma at time T16.

```{r}
EndovsRC.time<-emmeans(EndovsRC_m3, ~Tube|TimeInterval)
plot(EndovsRC.time, comparisons=TRUE)
```
```{r}
pairs(EndovsRC.time)
```



```{r}
# Pairwise comparisons
gene_pwc <- gene_level_ratios %>% group_by(Tube) %>% pairwise_t_test(EndovsRC ~ TimeInterval) %>% add_xy_position(x = "TimeInterval")

# Show p-values
# ggplot(gene_level_ratios, aes(x = TimeInterval, y = EndovsRC, color = TimeInterval)) +
#   geom_boxplot()+
  # geom_point()+
  # scale_colour_manual(values=color_panel)+
  # facet_wrap(~ Tube) +
  # stat_pvalue_manual(gene_pwc, label = "p.signif") +
  # labs(title = "Time lapse comparison within tubes (T test)", y = "miRNA vs RC concentration")
```



```{r, include=FALSE}
#Compare models
EndovsRC_m3_noInter<-lme(EndovsRC~Tube+RNAisolation+TimeInterval+Tube,
       random=~1|donorID, 
       data=gene_level_ratios, method = "ML")

EndovsRC_m3_Inter<-lme(EndovsRC~Tube+RNAisolation+TimeInterval+Tube:TimeInterval,
       random=~1|donorID, 
       data=gene_level_ratios, method = "ML")

EndovsRC_m3_allInter<-lme(EndovsRC~Tube*RNAisolation*TimeInterval,
       random=~1|donorID, 
       data=gene_level_ratios, method = "ML")

anova(EndovsRC_m3_noInter, EndovsRC_m3_Inter)
anova(EndovsRC_m3_noInter, EndovsRC_m3_allInter)
anova(EndovsRC_m3_Inter, EndovsRC_m3_allInter)
```



## RNA Extraction efficiency

Calculation: Endo/LP * EluateV/PlasmaInputV

```{r}
test <- log_rescaling_CI(gene_level_ratios, measurevar="EndovsLP_InputCorr", groupvar=c("GraphTube2"))
# Plot LP/endo in log10 scale
spikes_conc <- ggplot(test, aes(x=GraphTube2, y=10^measurevar_log_resc)) + 
  #geom_bar(position=position_dodge(), stat="identity")+
  facet_wrap(~RNAisolation, scales = "free_x", ncol = 2)+
  geom_errorbar(aes(ymin=ci_lower_oriscale, ymax=ci_upper_oriscale), colour="grey", width=.1) +
  geom_point(aes(colour=Tube), size=1.2) +
  geom_point(data=test, aes(x=GraphTube2, y=mean_oriscale), shape=4, colour="grey") +
  geom_hline(yintercept = 1, linetype="dashed",colour="grey88") +
  labs(x="", y="relative endogenous small RNA concentration", title=NULL, subtitle="endogenous small RNA vs LP (rescaled)", col = NULL) +
  scale_colour_manual(values=color_panel) +
  scale_y_log10() +
  theme_bar 

ggplotly(spikes_conc)
```

```{r}
EndovsLP_InputCorr_m1<-lme(EndovsLP_InputCorr~Tube*RNAisolation*TimeInterval,
       random=~1|donorID, 
       data=gene_level_ratios)

anova(EndovsLP_InputCorr_m1)
```
```{r}
anova_EndovsLP_InputCorr_m1 <- round(anova(EndovsLP_InputCorr_m1)[c(5, 6, 7), c(4)], digits = 3)
```


```{r}
EndovsLP_InputCorr_m2<-lme(EndovsLP_InputCorr~(Tube+RNAisolation+TimeInterval)^2,
       random=~1|donorID, 
       data=gene_level_ratios)

anova(EndovsLP_InputCorr_m2)
```
```{r}
EndovsLP_InputCorr_m3<-lme(EndovsLP_InputCorr~Tube+RNAisolation+TimeInterval+Tube:RNAisolation+Tube:TimeInterval,
       random= ~1|donorID, 
       data=gene_level_ratios)

anova(EndovsLP_InputCorr_m3)
```


```{r}
qqnorm(EndovsLP_InputCorr_m3$residuals)
qqline(EndovsLP_InputCorr_m3$residuals)
```
```{r}
EndovsLP_InputCorr_m.time<-emmeans(EndovsLP_InputCorr_m3, ~TimeInterval|Tube)
plot(EndovsLP_InputCorr_m.time, comparisons=TRUE)
```
```{r}
pairs(EndovsLP_InputCorr_m.time)
```
```{r}
EndovsLP_InputCorr_m.method<-emmeans(EndovsLP_InputCorr_m3, ~Tube|RNAisolation)
plot(EndovsLP_InputCorr_m.method, comparisons=TRUE)
```
```{r}
pairs(EndovsLP_InputCorr_m.method)
```

# Gene Count (Sensitivity)

number of miRNA with counts > 6

```{r}
cutoff_kit <- data.frame(Tube = sample_annotation$Tube, GraphTube = sample_annotation$GraphTube, median_th = 3)
```

```{r}
miRNAs_long <-  miRNAs %>% gather(., key="UniqueID", value="est_counts", -MIMATID) %>% #long format
  left_join(., dplyr::select(sample_annotation,c(UniqueID,RNAID,RNAisolation,GraphTube,GraphKit, SampleID, Tube, BiologicalReplicate)), by=c("UniqueID" = "RNAID"))
#keep only miRNA coding genes with more than 0 counts

miRNAs_long <-  miRNAs_long %>% filter(est_counts > 0)

number_miR_detected <-  miRNAs_long %>% group_by(SampleID) %>% dplyr::summarize(miR_above0=n()) 

number_miR_detected <- full_join(number_miR_detected, 
                                    miRNAs_long %>% group_by(SampleID) %>%
                                     dplyr::summarize(total_est_counts_above0=sum(est_counts)), 
                                   by="SampleID")
#cutoff_kit <- single_pos %>% group_by(GraphTube) %>% dplyr::summarise(median_th = median(filter_threshold))
 miRNAs_cutoff <-  miRNAs %>% gather(., key="UniqueID", value="est_counts", -MIMATID) %>% #long format
  left_join(., dplyr::select(sample_annotation,c(UniqueID,RNAID,RNAisolation,GraphTube,GraphKit, SampleID, Tube, BiologicalReplicate, TimeInterval)), by=c("UniqueID" = "RNAID")) 
 
  #left_join(., cutoff_kit, by="GraphTube") #add the median cut-off for each kit
 miRNAs_cutoff <-  miRNAs_cutoff %>% 
  filter(est_counts > 6)
 
number_miR_detected <- full_join(number_miR_detected, 
                                    miRNAs_cutoff %>% group_by(SampleID) %>%
                                     dplyr::summarize(miR_aboveTh=n()),
                                   by="SampleID")
number_miR_detected <- full_join(number_miR_detected, 
                                    miRNAs_cutoff %>% group_by(SampleID) %>%
                                     dplyr::summarize(total_est_counts_aboveTh=sum(est_counts)), 
                                   by="SampleID")
number_miR_detected <- left_join(number_miR_detected, 
                                   dplyr::select(sample_annotation, c(UniqueID,RNAID, GraphTube, GraphKit, RNAisolation, EluateV,SampleID, Tube, BiologicalReplicate, TimeInterval, donorID)),
                                   by="SampleID")
#convert to the original format
 miRNAs_cutoff <- dplyr::select(miRNAs_cutoff, MIMATID, UniqueID, est_counts, Tube, BiologicalReplicate, TimeInterval) %>% 
  spread(., key=UniqueID, value=est_counts)
#write.csv( miRNAs_cutoff, file="../../../exRNAQC/ miRNAs_cutoff.csv",row.names=FALSE, na="")
```

```{r}
count_m1<-lme(miR_aboveTh~Tube*RNAisolation*TimeInterval,
       random=~1|donorID, 
       data=number_miR_detected)

anova(count_m1)
```
```{r}
anova_count_m1 <- round(anova(count_m1)[c(5, 6, 7), c(4)], digits = 3)
```


```{r}
count_m2<-lme(miR_aboveTh~(Tube+RNAisolation+TimeInterval)^2,
       random=~1|donorID, 
       data=number_miR_detected)

anova(count_m2)
```
```{r}
count_m3<-lme(miR_aboveTh~Tube+RNAisolation+TimeInterval+Tube:RNAisolation+Tube:TimeInterval,
       random=~1|donorID, 
       data=number_miR_detected)

anova(count_m3)
```

```{r}
qqnorm(count_m3$residuals)
qqline(count_m3$residuals)
```
```{r}
count_m3.time<-emmeans(count_m3, ~TimeInterval|Tube)
plot(count_m3.time, comparisons=TRUE)
```
```{r}
pairs(count_m3.time)
```
```{r}
count_m3.time<-emmeans(count_m3, ~Tube|TimeInterval)
plot(count_m3.time, comparisons=TRUE)
```

```{r}
pairs(count_m3.time)
```

```{r}
#T test
# Pairwise comparisons
count_pwc <-  number_miR_detected %>% group_by(TimeInterval) %>% pairwise_t_test(miR_aboveTh ~ Tube) %>% add_xy_position(x = "Tube")


# Show p-values
count_plot<-ggplot(number_miR_detected, aes(x = Tube, y = miR_aboveTh, color=Tube)) +
  geom_boxplot() +
  geom_point() +
  facet_wrap(~ as.factor(TimeInterval)) +
  stat_pvalue_manual(count_pwc, label = "p.adj") +
  scale_color_manual(values=color_panel)+
  scale_fill_manual(values=color_panel)+
  theme_point + theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1), legend.position = 'none')+
  labs(y = "sensitivity")
count_plot

```




```{r}
count_m3.method<-emmeans(count_m3, ~Tube|RNAisolation)
plot(count_m3.method, comparisons=TRUE)
```


```{r}
pairs(count_m3.method)
```

```{r}
# Pairwise comparisons
count_pwc <-  number_miR_detected %>% group_by(GraphKit) %>% pairwise_t_test(miR_aboveTh ~ Tube) %>% add_xy_position(x = "Tube")


# Show p-values
count_plot<-ggplot(number_miR_detected, aes(x = Tube, y = miR_aboveTh, color=Tube)) +
  geom_boxplot() +
  geom_point() +
  facet_wrap(~ as.factor(GraphKit)) +
  stat_pvalue_manual(count_pwc, label = "p.adj", y.position = c(215, 225, 235)) +
  scale_color_manual(values=color_panel)+
  scale_fill_manual(values=color_panel)+
  theme_point + theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1), legend.position = 'none')+
  labs(y = "sensitivity")
count_plot

ggsave("./smallSensitivity.pdf", plot = count_plot, dpi = 300)
```


# Area left of the curve (ALC)

- The ALC (area-left-of-curve) calculation is done according to see Mestdagh et al., Nature Methods, 2014, and represents a precision metric:
  1. Only genes that reach threshold (95% single positive (SP) elimination based on exRNAQC004) are taken into account.
  2. All zero counts are replaced by NA
  3. The log2 ratios of counts for each gene are determined, each time between 2 timepoints
  4. Then, the absolute value of these log2 ratios are taken and ranked. This is plotted for every condition

- The faster the curve reaches 100% -> the smaller the ALC -> the better the precision (indicates that the replicates give similar counts for each gene)

The ALC values are summarized based on the average of the ALC values (after removing NAs). The dot plot at the end gives an overview of all these values.

**Score: lower ALC = better**

<!-- ## ALC - At least 1 replicate > threshold (and other > 0) -->

```{r, fig.height=2, fig.width=2, message=FALSE, warning=FALSE, echo=FALSE, fig.show="hold"}
double_positives<-miRNAs %>% dplyr::select(starts_with("RNA"),MIMATID) %>%
  replace(., is.na(.),0) %>% #first change NAs to 0
  mutate_if(is.numeric, funs(round)) #round everything to nearest integer
 
double_positives<-double_positives[rowSums(double_positives %>% select_if(is.numeric) > 1, na.rm=T)>0,] # keep only the rows where at least one sample has more than 1

generate_ALC <- function(sample_annotation, filter_key, filter_val){
  sample_annotation <- sample_annotation %>%
    filter((!!as.symbol(filter_key)) == filter_val)
  ALC_input_all <- data.frame() 
  for (duplicate_type in unique(sample_annotation$Tube)){
    sample_duplicates<-sample_annotation%>% filter(Tube==duplicate_type) %>%
      dplyr::select(UniqueID,RNAID,RNAisolation,GraphTube,TimeInterval,Tube, BiologicalReplicate)
    cutoff95SP <- cutoff_kit[cutoff_kit$Tube==duplicate_type,]$median_th
    if(nrow(sample_duplicates)>1){
      #print(duplicate_type)
      double_positives_sample<-double_positives %>% dplyr::select(MIMATID,sample_duplicates$RNAID) # only keep the replicates of one type
      samples_comb <- combn(sample_duplicates$RNAID,2) #compare 2 of the 3 samples at a time
      for (n_col in 1:ncol(samples_comb)) {
        #print(samples_comb[,n_col])
        nr_runA <- gsub("^[A-Z]+","",sample_annotation[sample_annotation$RNAID== samples_comb[1,n_col],]$TimeInterval)
        nr_runB <- gsub("^[A-Z]+","",sample_annotation[sample_annotation$RNAID== samples_comb[2,n_col],]$TimeInterval)
        RepA <- sample_annotation[sample_annotation$RNAID== samples_comb[1,n_col],]$Donor
        RepB <- sample_annotation[sample_annotation$RNAID== samples_comb[2,n_col],]$Donor
       if ((RepA == RepB)){
        varname <- paste0("T",nr_runA,"_",nr_runB) #make a name so you can backtrace which replicates are compared
        #print(varname)
        double_positives_2repl <- double_positives %>% dplyr::select(MIMATID, paste(samples_comb[1,n_col]), paste0(samples_comb[2,n_col])) %>%
          filter_if(is.numeric, all_vars(.>0)) %>% #keep only the 2 replicates of one kit and keep only genes where both replicates have > 0 counts
          filter_if(is.numeric, any_vars(.>cutoff95SP)) # remove genes where neither of the replicates reach the threshold that removes 95% of SP in that kit
        correlation_samples<-double_positives_2repl %>%
          mutate(log2_ratio=abs(log(get(samples_comb[1,n_col]),2)-log(get(samples_comb[2,n_col]),2))) %>%
          dplyr::select(log2_ratio,MIMATID) #%>% drop_na()
        ALC_input<-correlation_samples %>% arrange(log2_ratio) %>% # order by log2 ratio and then make a rank (counter) and rescale this to 1 (perc_counter)
          #mutate(rank=percent_rank(log2_ratio)) %>% # this does not work: gives everything with log2ratio = 0 rank 0
          mutate(counter = seq(1:nrow(double_positives_2repl)), Tube=duplicate_type, Replicates=varname, BiologicalReplicate = paste0(RepA,"-",RepB)) %>%
          mutate(ReplID = paste0(Tube,"-",Replicates), perc_counter = counter/nrow(double_positives_2repl))
       
        ALC_input_all <- rbind(ALC_input_all, ALC_input)
        }
      }
    }
  }
  ALC_input_all[filter_key] <- filter_val
  return(ALC_input_all)
}

ALC_input_all <- data.frame() 
for (method in unique(sample_annotation$RNAisolation)){
  temp <-generate_ALC(sample_annotation, filter_key = 'RNAisolation', filter_val = method)
  ALC_input_all <- rbind(ALC_input_all, temp)
}
max_ALC <-max(ALC_input_all$log2_ratio) # calculate the max ALC over everything (necessary for area calculation -> should always be the same in order to compare among kits)

ALC <- ALC_input_all %>% dplyr::group_by(Tube,Replicates,BiologicalReplicate, RNAisolation) %>%
  dplyr::summarise(ALC_calc = sum(log2_ratio)/(max_ALC*length(MIMATID))) %>%
  #dplyr::summarise(ALC_calc = sum(log2_ratio)/(max_ALC)) %>%
  mutate(ReplID = paste0(Tube,"-",Replicates))




ALC_melt <- left_join(ALC, sample_annotation, by="Tube")
ALC$BiologicalReplicate <- gsub("[-]TUBE[0-9]*","", ALC$BiologicalReplicate)
ALC$Replicates <- gsub("^Rep04_0$","T0-T04", ALC$Replicates)
ALC$Replicates <- gsub("^Rep0_04$","T0-T04", ALC$Replicates)
ALC$Replicates <- gsub("^Rep16_0$","T0-T16", ALC$Replicates)
ALC$Replicates <- gsub("^Rep0_16$","T0-T16", ALC$Replicates)
ALC$Replicates <- gsub("^Rep16_04$","T04-T16", ALC$Replicates)
ALC$Replicates <- gsub("^Rep04_16$","T04-T16", ALC$Replicates)
ALC$Replicates <- gsub("^Rep24_0$","T0-T24", ALC$Replicates)
ALC$Replicates <- gsub("^Rep0_24$","T0-T24", ALC$Replicates)
ALC$Replicates <- gsub("^Rep72_0$","T0-T72", ALC$Replicates)
ALC$Replicates <- gsub("^Rep0_72$","T0-T72", ALC$Replicates)
ALC$Replicates <- gsub("^Rep72_24$","T24-T72", ALC$Replicates)
ALC$Replicates <- gsub("^Rep24_72$","T24-T72", ALC$Replicates)
#ALC_melt <- ALC_melt %>% filter(Replicates %in% c("Rep0_04", "Rep04_16", "Rep0_16", "Rep0_24", "Rep24_72", "Rep0_72"))

ALC$TimePoint <- ifelse(ALC$Replicates %in% c("T24_0", "T0_24", "T0_04", "T04_0"), "1st timepoint", ifelse(ALC$Replicates%in% c("T72_0", "T0_72", "T0_16", "T16_0"), "2nd timepoint", NA))
ALC$TubeType <- ifelse(ALC$Tube %in% c("DNA Streck", "Biomatrica", "RNA Streck", "PAXgene", "Roche"), "preservation", "non-preservation")

ALC_mean <- ALC %>% dplyr::group_by(Tube) %>% dplyr::summarise(ALC = mean(ALC_calc))

ALC<-ALC %>% filter(!(is.na(TimePoint)))
```




```{r}
ALC_m1<-lme(ALC_calc~Tube*RNAisolation*TimePoint,
       random=~1|BiologicalReplicate, 
       data=ALC)

anova(ALC_m1)
```
```{r}
anova_ALC_m1 <- round(anova(ALC_m1)[c(5, 6, 7), c(4)], digits = 3)
```

There is a significant interaction between Tube:RNAIsolatiom

Now, second-order interactions

```{r}
ALC_m2<-lme(ALC_calc~(Tube+RNAisolation+TimePoint)^2,
       random=~1|BiologicalReplicate, 
       data=ALC)

anova(ALC_m2)
```

There interaction between Tube:RNAIsolatiom remains significant.

Now, the model keeping only the significant interaction

```{r}
ALC_m3<-lme(ALC_calc~Tube+RNAisolation+TimePoint+Tube:RNAisolation,
       random=~1|BiologicalReplicate, 
       data=ALC)

anova(ALC_m3)
```

```{r}
qqnorm(ALC_m3$residuals)
qqline(ALC_m3$residuals)
```

```{r}
ALC_m3.methods<-emmeans(ALC_m3, ~Tube|RNAisolation)
plot(ALC_m3.methods, comparisons=TRUE)
```
```{r}
pairs(ALC_m3.methods)
```

```{r}
# Pairwise comparisons
ALC_pwc <- ALC %>% group_by(RNAisolation) %>% pairwise_t_test(ALC_calc ~ Tube) %>% add_xy_position(x = "Tube")

# Show p-values
alc_plot<-ggplot(ALC, aes(x = Tube, y = ALC_calc, color = Tube)) +
  geom_boxplot()+
  geom_point()+
  scale_colour_manual(values=color_panel)+
  facet_wrap(~ RNAisolation) +
  stat_pvalue_manual(ALC_pwc, label = "p.signif") +
  theme(panel.grid.major.x=element_line(linetype = "dashed",color="gray88"), axis.text.x=element_text(size=8))+
  labs(title = "Tube comparison within kits (T test)", y = "Area left of the curve")
alc_plot
```


# Biotype: Percentage of miRNAs counts

We are usually interested in miRNAs, and the other biotypes are often seen as contaminants (see above). The fraction was calculated as the number of miRNA reads / total reads (mapped + spikes)

```{r, message=FALSE, warning=F}
perc_miRs <- miRNAs %>% gather(key="UniqueID",value="counts", -"MIMATID") %>% mutate(type=gsub("[0-9].*$","",MIMATID)) %>%
  dplyr::group_by(UniqueID,type) %>% 
  dplyr::summarise(sum_counts = sum(counts, na.rm=T)) %>% #calculate sum of MIMAT per sample
  full_join(reads_complete %>% filter(reads=="all_mapped")) %>% 
  mutate(perc=sum_counts/counts*100) %>% #calculate perc of miRs and spikes of total mapped reads
  left_join(sample_annotation, by=c("UniqueID" = "RNAID"))
```

```{r}
Biotype_m1<-lme(perc~Tube*RNAisolation*TimeInterval,
       random=~1|BiologicalReplicate, 
       data=perc_miRs)

anova(Biotype_m1)
```

```{r}
anova_Biotype_m1 <- round(anova(Biotype_m1)[c(5, 6, 7), c(4)], digits = 3)
```

Second order:

```{r}
Biotype_m2<-lme(perc~(Tube+RNAisolation+TimeInterval)^2,
       random=~1|BiologicalReplicate, 
       data=perc_miRs)

anova(Biotype_m2)
```

Keeping  only the significant interactions:

```{r}
Biotype_m3<-lme(perc~Tube+RNAisolation+TimeInterval+Tube:RNAisolation+Tube:TimeInterval,
       random=~1|BiologicalReplicate, 
       data=perc_miRs)
anova(Biotype_m3)
```

```{r}
Biotype_m3.methods<-emmeans(Biotype_m3, ~Tube|RNAisolation)
plot(Biotype_m3.methods, comparisons=TRUE)
```

```{r}
pairs(Biotype_m3.methods)
```

```{r}
# Pairwise comparisons
Biotype_pwc <- perc_miRs %>% group_by(RNAisolation) %>% pairwise_t_test(perc ~ Tube) %>% add_xy_position(x = "Tube")

# Show p-values
Biotype_plot<-ggplot(perc_miRs, aes(x = Tube, y = perc, color = Tube)) +
  geom_boxplot()+
  geom_point()+
  scale_colour_manual(values=color_panel)+
  facet_wrap(~ RNAisolation) +
  stat_pvalue_manual(Biotype_pwc, label = "p.signif") +
  theme(panel.grid.major.x=element_line(linetype = "dashed",color="gray88"), axis.text.x=element_text(size=8))+
  labs(title = "Tube comparison within kits (T test)", y = "miRNA counts (%)")
Biotype_plot
```

```{r}
Biotype_m3.time<-emmeans(Biotype_m3, ~TimeInterval|Tube)
plot(Biotype_m3.time, comparisons=TRUE)
```
```{r}
pairs(Biotype_m3.time)
```


```{r}
Biotype_m3.time<-emmeans(Biotype_m3, ~Tube|TimeInterval)
plot(Biotype_m3.time, comparisons=TRUE)
```
```{r}
pairs(Biotype_m3.time)
```

# Conclusion

```{r}
overview <- data.frame(row.names = c("Tube:RNAisolation","Tube:TimeInterval","RNAisolation:TimeInterval"), "RNA concentration" = anova_EndovsRC_m1, "extraction efficiency" = anova_EndovsLP_InputCorr_m1, "sensitivity" = anova_count_m1, "reproducibility" = anova_ALC_m1, "biotype" = anova_Biotype_m1, check.names = FALSE)
overview_t <- t(overview)

#breaks for heatmap
bk1 <- c(seq(0,0.051,by=0.005))

#make color palette
custom_palette = c(colorRampPalette(brewer.pal(n = 7, name =
  "RdBu"))(20))
custom_palette2 = c(custom_palette[1:10], 'white')

#make matrix with labels for heatmap
labels_tab <- overview_t
labels_tab[labels_tab<0.001]<-"<0.001"

#make heatmap
phm<-pheatmap::pheatmap(overview_t,
                   breaks = bk1,
                   display_numbers = labels_tab,fontsize_number=11, number_color="black",
                   color = custom_palette2,
                   angle_col = 45,
                   cluster_rows=FALSE, cluster_cols=FALSE)

ggsave("./smallInteractions2.pdf", plot = phm, dpi = 300)
```


